Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-02-25 main.Rmd
nonlinear to cyclic terminology - consistent with mid_can review document

---
title: Visualizing probability distributions across bivariate nonlinear temporal granularities

@@

abstract: |
 Recent advances in technology greatly facilitates recording and storing data at much finer temporal scales than was previously possible. As the frequency of time-oriented data increases, the number of questions about the observed variable that can be addressed by visual representation also increases. We propose some new tools to explore this type of data, which deconstruct time in many different ways. There are several classes of time deconstructions including linear and nonlinear time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. Nonlinear time granularities can be circular such as hour of the day, and day of the week, quasi-circular such as day of the month, and aperiodic such as public or school holidays.

@@

Linear time granularities respect the linear progression of time such as hours, days, weeks and months, and non-repeating. One of the first attempts to characterize these granularities is due to @Bettini1998-ed. However, the definitions and rules defined are inadequate for describing nonlinear or repeating granularities. Hence, there is a need to define some new nonlinear time granularities, that can be useful in visualizations. Nonlinear time granularities can be circular, quasi-circular or aperiodic. Examples of circular granularities are hour of the day and day of the week; an examples of a quasi-circular granularity is day of the month; examples of aperiodic granularities are public holidays and school holidays.

@@

  * provide a formal characterization of nonlinear granularities;
  * facilitate manipulation of single- and multiple-order-up time granularities through cyclic calendar algebra;
  * develop an approach to check the feasibility of creating plots or drawing inferences for any two nonlinear granularities;
  * recommend prospective probability distributions for exploring distributions of a univariate dependent variable across pair of granularities.

@@

The remainder of the paper is organized as follows: Section \ref{sec:linear-time} provide some background material on linear granularities and introduces calendar algebra for computing different linear granularities. Section \ref{sec:cyclic-gran} formally characterizes different nonlinear time granularities by extending the framework of linear time granularities. Section \ref{sec:cyclic-calendar} introduces calendar algebra for computing nonlinear time granularities. Section \ref{sec:data-structure} discusses the data structure for exploring the conditional distributions of the associated time series across pairs of nonlinear time granularities. Section \ref{sec:visualization} discusses the role of different factors in constructing an informative and trustworthy visualization. Section \ref{sec:application} examines how systematic exploration can be carried out for a temporal and non-temporal application. Section \ref{sec:discussion} summarizes this paper and discusses possible future direction.

@@

# Nonlinear time granularities {#sec:cyclic-gran}

@@

We propose a formalism of nonlinear time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. In a tsibble, each observation (row) is uniquely identified by index and key.

@@

Nonlinear time granularities with periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular (minutes and hours) or irregular (days and months). <!--For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31).--> A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping. Nonlinear time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mapping is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.

@@

Aperiodic nonlinear granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$, that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity. Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to Definition \ref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$ and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

\begin{definition}[Aperiodic nonlinear granularity]\label{aperiodic}
An aperiodic nonlinear granularity $A_{B, M}$ relates aperiodic linear granularity $M$ and bottom granularity $B$, if
\begin{equation}\label{eq:aperiodic}
\begin{split}
A_{B, M}(z) & = L(i), \quad z \in T_{i_j} \quad and \quad \forall i \in {1, 2, \dots, n}\\
     & = L(0), \quad otherwise
\end{split}
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$T_{i_j}$ are the sets of indices of $B$ describing aperiodic linear granularities $M_{i}(j)$ such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$, $M = \bigcup_{i=1}^{n}M_{i}$.
\end{definition}

@@

```{r aperiodic-example, echo=FALSE,out.width = "100%",fig.cap="Aperiodic nonlinear granularity $A(B, M1)$ relating aperiodic linear granularity $M1$ and bottom granularity $B$"}
knitr::include_graphics("Figs/aperiodic-example.png")
```

@@

In \autoref{sec:cyclic-gran}, we discussed how we can define nonlinear granularities by extending the framework of linear granularities defined in @Bettini1998-ed. In this section, we will see how to obtain cyclic granularities through an algebraic representation of other cyclic granularities. This is similar to calendar algebra in @Ning2002-tf, where linear granularities are generated from the bottom linear granularity or from those that are already generated. Since, our method caters to computation of cyclic granularities, we shall refer to it as "cyclic calendar algebra".

>>>>

---
title: Visualizing probability distributions across bivariate cyclic temporal granularities

@@

abstract: |
 Recent advances in technology greatly facilitates recording and storing data at much finer temporal scales than was previously possible. As the frequency of time-oriented data increases, the number of questions about the observed variable that can be addressed by visual representation also increases. We propose some new tools to explore this type of data, which deconstruct time in many different ways. There are several classes of time deconstructions including linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and monthsCyclic time granularities can be circular such as hour of the day, and day of the week, quasi-circular such as day of the month, and aperiodic such as public or school holidays.

@@

Linear time granularities respect the linear progression of time such as hours, days, weeks and months, and non-repeating. One of the first attempts to characterize these granularities is due to @Bettini1998-ed. However, the definitions and rules defined are inadequate for describing cyclic or repeating granularities. Hence, there is a need to define some new cyclic time granularities, that can be useful in visualizationsCyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular granularities are hour of the day and day of the week; an examples of a quasi-circular granularity is day of the month; examples of aperiodic granularities are public holidays and school holidays.

@@

  * provide a formal characterization of cyclic granularities;
  * facilitate manipulation of single- and multiple-order-up time granularities through cyclic calendar algebra;
  * develop an approach to check the feasibility of creating plots or drawing inferences for any two cyclic granularities;
  * recommend prospective probability distributions for exploring distributions of a univariate dependent variable across pair of granularities.

@@

The remainder of the paper is organized as follows: Section \ref{sec:linear-time} provide some background material on linear granularities and introduces calendar algebra for computing different linear granularities. Section \ref{sec:cyclic-gran} formally characterizes different cyclic time granularities by extending the framework of linear time granularities. Section \ref{sec:cyclic-calendar} introduces calendar algebra for computing cyclic time granularities. Section \ref{sec:data-structure} discusses the data structure for exploring the conditional distributions of the associated time series across pairs of cyclic time granularities. Section \ref{sec:visualization} discusses the role of different factors in constructing an informative and trustworthy visualization. Section \ref{sec:application} examines how systematic exploration can be carried out for a temporal and non-temporal application. Section \ref{sec:discussion} summarizes this paper and discusses possible future direction.

@@

# cyclic time granularities {#sec:cyclic-gran}

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. In a tsibble, each observation (row) is uniquely identified by index and key.

@@

cyclic time granularities with periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular (minutes and hours) or irregular (days and months). <!--For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31).--> A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping. Cyclic time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mapping is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.

@@

Aperiodic cyclic granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$, that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity. Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to Definition \ref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$ and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

\begin{definition}[Aperiodic cyclic granularity]\label{aperiodic}
An aperiodic cyclic granularity $A_{B, M}$ relates aperiodic linear granularity $M$ and bottom granularity $B$, if
\begin{equation}\label{eq:aperiodic}
\begin{split}
A_{B, M}(z) & = L(i), \quad z \in T_{i_j} \quad and \quad \forall i \in {1, 2, \dots, n}\\
     & = L(0), \quad otherwise
\end{split}
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$T_{i_j}$ are the sets of indices of $B$ describing aperiodic linear granularities $M_{i}(j)$ such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$, $M = \bigcup_{i=1}^{n}M_{i}$.
\end{definition}

@@

```{r aperiodic-example, echo=FALSE,out.width = "100%",fig.cap="Aperiodic cyclic granularity $A(B, M1)$ relating aperiodic linear granularity $M1$ and bottom granularity $B$"}
knitr::include_graphics("Figs/aperiodic-example.png")
```

@@

In \autoref{sec:cyclic-gran}, we discussed how we can define cyclic granularities by extending the framework of linear granularities defined in @Bettini1998-ed. In this section, we will see how to obtain cyclic granularities through an algebraic representation of other cyclic granularities. This is similar to calendar algebra in @Ning2002-tf, where linear granularities are generated from the bottom linear granularity or from those that are already generated. Since, our method caters to computation of cyclic granularities, we shall refer to it as "cyclic calendar algebra".
